Abstractions for Multi-Sorted Substitutions

### 2023-09-19 paper/paper.tex
Update.

  % A \ACode{Kit} is a structure that captures the commonalities
  % between variables and terms:
  A \ACode{Kit} is a structure that allows to abstract over whether
  something is a term or a variable. The intention is to instanciate
  this structure exactly twice (once for variables and once for terms),
  and then write definitions, which are parameterized over a kit and
  consequently can be used for both variables and terms.
  \LibCode\KKit

  As we intend to have exactly two \ACode{\ACon{Kit}} instances, we choose names of
  the form \ACode{x/y}, where \ACode{x} is the name we choose for the
  variable instance, and \ACode{y} is the name we choose for the term
  instance.
  For example the parameter type \ACode{\_∋/⊢\_} will be instanciated
  to \ACode{\ACon{\_∋\_}} for the variable kit, and to \ACode{\AField{\_⊢\_}} for the
  term kit.

  A kit provides the following operations:
  \begin{itemize}
  \item 
    \ACode{\AField{id/`}} converts a variable \ACode{S \ACon{∋} s} into a
    \ACode{S \AField{∋/⊢} s}. For the variable kit, \ACode{S ∋/⊢ s} is instanciated to \ACode{S \ACon{∋} s}, so this
    operation is the identity. For the term kit, \ACode{S ∋/⊢ s} is
    instanciated to \ACode{S \AField{⊢} s}, so this operation is the
    variable constructor \ACode{\AField{`\_}}.
  \item 
    \ACode{\AField{`/id}} converts a \ACode{S ∋/⊢ s} into a term \ACode{S \AField{⊢} s}
    and is dual to the \ACode{\AField{id/`}} operation.
  \item 
    \ACode{\AField{wk}} shifts the DeBruin-Indices in a \ACode{S ∋/⊢ s}.
    The new, unused variable \ACode{\ACon{zero}} can assume any sort \ACode{s'}.
    For variables, \ACode{\AField{wk}} is the successor \ACode{\ACon{suc}}.
    For terms, \ACode{\AField{wk}} means applying a shifting renaming to the term.
  \end{itemize}

@@

  Note that all definitions so far are library code. To make them
  available for our System F example, all we need to do is
  to create an instance of \ACode{\ACon{Traversal}}.

  We start with the operation of applying a map to a term:
  \AppCode\FTraversalOp
  The interesting cases are those with variables and binders:
  \begin{itemize}
  \item 
    In the variable case \ACode{(\ACon{`} x) \ADef{⋯} ϕ}, we first apply the map
    \ACode{ϕ} to the variable \ACode{x}. If \ACode{ϕ} is a renaming,
    we get back a variable and need to apply the variable constructor
    \ACode{\ACon{`\_}} to it. If \ACode{ϕ} is a substitution,
    we get back a term that we can use directly.
    This is exactly what \ACode{\AField{`/id}} does.
  \item 
    In cases where the operation needs to go under a binder, like
    \ACode{(\ACon{λx} e) \ADef{⋯} ϕ}, we lift the map using \AField{\_↑\_} before
    we apply it to the subterm to account for the bound variable.
  \end{itemize}

@@

  Finally, we instanciate the \ACode{\ACon{Traversal}} structure:
  % \LibCode\FTraversalIdProofRest
  \AppCode\FTraversal

@@

  This prevented us to define a useful lemma, which also needs to
  be instanciated separately for variables and terms: the counterpiece
  to \ACode{\AField{wk-id/`}}, which we call \ACode{\AField{wk-`/id}}.

@@

  Using the \ACode{\ACon{WkKit}}, we then define
  the \ACode{\ACon{ComposeKit}} and \ACode{\ACon{ComposeTraversal}}
  structures, which serve analogous purposes to \ACode{\ACon{Kit}}
  and \ACode{\ACon{Traversal}} from before:
  the library provides a \ACode{\ACon{ComposeKit}} for each of the
  four compositions, whereas the \ACode{\ACon{ComposeTraversal}}
  structure needs to be instanciated by the application with
  lemmas parameterized over arbitrary \ACode{\ACon{ComposeKit}}s.
  This gives the user a composition operation
  \ACode{\AField{\_·ₖ\_}} between arbitrary maps and corresponding
  lemmas, instead of multiple definitions and lemmas about concrete
  maps, enabling us to reason on a more abstract level and to avoid
  further boilerplate.

@@

  Similar as for the \ACode{\ACon{Traversal}}, we derive further
  definitions from the the axiom of a \ACode{\ACon{ComposeTraversal}}:
  \begin{itemize}
  \item 
    A map \ACode{ϕ} followed by a weakening is equivalent to a
    weakening followed by \ACode{ϕ} that has been lifted over the weakened
    variable:
    \LibCode\KCommLiftWeaken
    % \LibCode\KCommLiftWeakenProof
    By applying \ACode{\AField{⋯-fusion}} to this lemma, we get the same
    statement for multiple map applications:
    \LibCode\KCommLiftWeakenTraverse
    % \LibCode\KCommLiftWeakenTraverseProof
  \item 
    A weakening followed by a singleton substitution cancel each other out:
    \LibCode\KWeakenCancelsSingle
    % \LibCode\KWeakenCancelsSingleProof
    By applying \ACode{\AField{⋯-fusion}} and \ACode{\AField{⋯-id}} to
    this lemma, we get the same statement for multiple map applications:
    \LibCode\KWeakenCancelsSingleTraverse
    % \LibCode\KWeakenCancelsSingleTraverseProof
  \item 
    A singleton substitution followed by some map \ACode{ϕ} is the same
    as \ACode{ϕ} lifted over the variable of the singleton substitution
    followed by the singleton substitution with \ACode{ϕ} applied to
    its content:
    \LibCode\KDistLiftSingle
    % \LibCode\KDistLiftSingleProof
    By applying \ACode{\AField{⋯-fusion}} to this lemma, we get the
    same statement for multiple map applications:
    \LibCode\KWeakenCancelsSingleTraverse
    \LibCode\KDistLiftSingleTraverse
    % \LibCode\KDistLiftSingleTraverseProof
  \item 
    Given the \ACode{\ADef{⋯-↑-wk}} lemma, we can define the four
    \ACode{\ACon{ComposeKit}} instances:
    \LibCode\KComposeKitInstances
    \LibCode\KComposeKitInstancesConcrete
  \end{itemize}

  Finally, we instantiate the \ACode{\ACon{ComposeTraversal}} for
  System F to make the composition machinery available:
  \AppCode\FFusion
  \AppCode\FFusionProofInteresting
  % \AppCode\FFusionProofRest
  \AppCode\FComposeTraversal

@@

  In the context of multi-sorted syntax, the notation of a type can be
  described as a mapping between sorts.
  For System F, the expression sort \ACode{\ACon{𝕖}} maps to the type sort
  \ACode{\ACon{𝕥}} and the sort \ACode{\ACon{𝕥}} maps to the kind sort
  \ACode{\ACon{𝕜}}.
  To teach our framework about types, we have to instantiate the following structure:
  \LibCode\KTypeSorts
  For System F the instanciation is
  \AppCode\FTypes
  For a nicer interface, we define a type analogously to a term:
  \LibCode\KTypes

@@

  \section{Typing}
  \label{sec:typing}
  Equipped with a notion of types and type context we are ready to define
  the multi-sorted typing relation for System F, incorporating both
  typing and kinding:
  \AppCode\FTyping
  The interesting cases are:
  \begin{itemize}
  \item
    the variable typing constructor covers both expression- and
    type-variables, analogously to the variable term constructor;
  \item
    the typing of the lambda abstraction \ACode{\ACon{⊢λ}} weakens the
    type of the codomain \ACode{t₂}.
    This is necessary, because with multi-sorted syntax, types could
    also depend on expressions, so the sub typing derivation accounts for
    another variable which is not used by the type.
  \item
    the kinding constructor \ACode{\ACon{⊢τ}} states that all types
    have kind \ACode{\ACon{★}}. This is sufficient for System F as types
    are automatically well-scoped due to intrinsic scoping.
  \end{itemize}
  To teach the framework about typing, we create a structure analogously to
  \ACode{\ACon{Syntax}}:
  \LibCode\KTyping
  and instanciate it for System F:
  \AppCode\FTypingInst

@@

  Given a \ACode{\ACon{TypingTraversal}}, we can easily instanciate the
  \ACode{\ACon{TypingKits}} for variables and terms:
  \LibCode\KTypingInstances
  % \LibCode\KTypingTraversalNotation

  We are then ready to create a \ACode{\ACon{TypingTraversal}}
  instance for System F:
  \AppCode\FPreserve
  \AppCode\FTypingTraversal

  As the \ACode{\ACon{Traversal}} and  \ACode{\ACon{ComposeTraversal}} instances
  can be generally derived via reflection,
  the \ACode{\AField{\_⊢⋯\_}} lemma is in fact the only lemma that
  has to been proved manually for subject reduction.
  From this single lemma follows type preservation for all operations
  of both renamings and substitutions for all sorts of variables
  simultaneously.

@@

  \section{Reflection \& Generics}
  \label{sec:reflection}
  We used Agda's reflection mechanism\cite{TODO} to derive
  instanciations related to all structures for untyped substitition,
  i.e. \ACode{\ACon{Traversal}} and \ACode{\ACon{ComposeTraversal}}.

  To gain insight into the class of object languages supported by our
  reflection algorithm, we modeled a class of object languages using a
  generic programming technique and instanciated the structures
  generically. The class contains all object languages which:
  \begin{enumerate}
  \item Have a variable constructor of type TODO;
  \item Uses subterms only directly (e.g. not in lists);
  \item Subterms only extend the scope-context.
  \end{enumerate}

@@

  \section{Case Studies}
  \label{sec:casestudies}
  \begin{itemize}
  \item
    For lambda calculus with dependent function types, the framework
    just works out of the box for both deriving and instanciating the
    \ACode{\ACon{TypingTraversal}}. Additionally, having the
    \ACode{\ACon{ComposeKit}} abstraction turned out to be useful
    to prove confluence, allowing to unify certain lemmas, which
    otherwise need to be proved for both renamings and substitutions,
    e.g. that for two terms \ACode{t} and \ACode{t'}
    with \ACode{t \ACon{↪} t'} it follows that \ACode{(t \AField{⋯} ϕ)
    \ACon{↪} (t' \AField{⋯} ϕ)} for any map \ACode{ϕ}.
  \item
    For System F with subtyping the main challenges are how to represent
    subtyping constraints and how to deal with the fact that
    substitution preserves typing is not generally true, as type variables
    have subtyping bounds that need to be respected.
    Instead of defining typing contexts and type preservation lemmas independent
    of our framework, we found a way that still allows us to benefit
    from the typing machinery.
    When binding a type variable with a subtyping constraint \ACode{α <: t},
    we first bind the type variable as \ACode{α : ★}, and then bind the constraint
    with a dummy name \ACode{\_ ∶ (α <: t)}. This description is
    similar to first-class constraints, but restricted enough to be
    isomorphic to the original formalization.
    With that encoding substitution preserves typing is generally true again, enabling
    the use of the framework's typing contexts and \ACode{\ACon{TypingKit}} machinery.
    Replacing a type variable \ACode{α <: t} with a type \ACode{t'}
    which is not a subtype of \ACode{t'}, the resulting term is still
    well-typed, but in a context with an insatisfiable constraint.
  \item
    Object language with pattern matching pose an interesting problem.
    A clause \ACode{p ⇒ e} in a match-expression consists of a pattern
    \ACode{p} and a continuation expression \ACode{e}.
    The pattern \ACode{p} can bind arbitrary many variables in subpatterns,
    which are then available as free variables in \ACode{e}.
    % The challenge is to track the sorts of variables bound by a pattern,
    % to make them available to \ACode{e} by extending \ACode{e}'s sort
    % context correspondingly.
    To track the sorts of variables bound by a pattern, one can define
    the sort of patterns to contain the sorts of the variables that it binds:
    % 𝕡 : List (Sort Var) → Sort NoVar
    The syntax for variable patterns and clauses can then be specified as:
    % varpat : S ⊢ 𝕡 [ 𝕖 ]  -- pattern binding an expression variable
    % pairpat : S ⊢ 𝕡 S₁ → S ⊢ 𝕡 S₂ → S ⊢ 𝕡 (S₁ ++ S₂)
    % clause : S ⊢ 𝕡 S' → (S' ++ S) ⊢ 𝕖 → S ⊢ 𝕔
  \end{itemize}

@@

  Allais et al.\cite{DBLP:conf/cpp/Allais0MM17} propose a powerful
  abstraction for denotational semantics and semantic fusion lemmas.
  In later work\cite{DBLP:journals/pacmpl/AllaisA0MM18}, they use
  generic programming to instanciate this abstraction for a class of
  object languages comparable to ours.
  They demonstrate how both renamings and substitutions can be
  described as semantics and how the four composition lemmas follow from
  their generic fusion lemma.
  They do cover an abstraction to unify renamings and substitutions.
  They do not show how to work with multiple variable sorts, how to
  abstract over the compositions, and how to abstract over typing in
  an extrinsically typed scenario.
  However, we believe that their framework is general enough such that
  our concepts can also be transferred to their framework.

>>>>

  % A \ACode{Kit} is a structure that captures the commonalities
  % between variables and terms:
  A \ACode{Kit} is a structure that allows to abstract over whether
  something is a term or a variable. The intention is to instantiate
  this structure exactly twice (once for variables and once for terms),
  and then write definitions, which are parameterized over a kit and
  consequently can be used for both variables and terms.
  \LibCode\KKit

  As we intend to have exactly two \ACode{\ACon{Kit}} instances, we choose names of
  the form \ACode{x/y}, where \ACode{x} is the name we choose for the
  variable instance, and \ACode{y} is the name we choose for the term
  instance.
  For example the parameter type \ACode{\_∋/⊢\_} will be instantiated
  to \ACode{\ACon{\_∋\_}} for the variable kit, and to \ACode{\AField{\_⊢\_}} for the
  term kit.

  A kit provides the following operations:
  \begin{itemize}
  \item 
    \ACode{\AField{id/`}} converts a variable \ACode{S \ACon{∋} s} into a
    \ACode{S \AField{∋/⊢} s}. For the variable kit, \ACode{S ∋/⊢ s} is instantiated to \ACode{S \ACon{∋} s}, so this
    operation is the identity. For the term kit, \ACode{S ∋/⊢ s} is
    instantiated to \ACode{S \AField{⊢} s}, so this operation is the
    variable constructor \ACode{\AField{`\_}}.
  \item 
    \ACode{\AField{`/id}} converts a \ACode{S ∋/⊢ s} into a term \ACode{S \AField{⊢} s}
    and is dual to the \ACode{\AField{id/`}} operation.
  \item 
    \ACode{\AField{wk}} shifts the DeBruin-Indices in a \ACode{S ∋/⊢ s}.
    The new, unused variable \ACode{\ACon{zero}} can assume any sort \ACode{s'}.
    For variables, \ACode{\AField{wk}} is the successor \ACode{\ACon{suc}}.
    For terms, \ACode{\AField{wk}} means applying a shifting renaming to the term.
  \end{itemize}

@@

  \subsection{Instantiation for System F}
  In this subsection, we show how to instantiate the
  \ACode{\ACon{Traversal}} abstraction for System F.
  In practice, this is done by our reflection algorithm automatically,
  but it can be instructive to see what happens under the hood.

  First, we define the operation of applying a map to a term:
  \AppCode\FTraversalOp
  The interesting cases are those with variables and binders:
  \begin{itemize}
  \item 
    In the variable case \ACode{(\ACon{`} x) \ADef{⋯} ϕ}, we first apply the map
    \ACode{ϕ} to the variable \ACode{x}. If \ACode{ϕ} is a renaming,
    we get back a variable and need to apply the variable constructor
    \ACode{\ACon{`\_}} to it. If \ACode{ϕ} is a substitution,
    we get back a term that we can use directly.
    This is exactly what \ACode{\AField{`/id}} does.
  \item 
    In cases where the operation needs to go under a binder, like
    \ACode{(\ACon{λx} e) \ADef{⋯} ϕ}, we lift the map using \AField{\_↑\_} before
    we apply it to the subterm to account for the bound variable.
  \end{itemize}

@@

  Finally, we instantiate the \ACode{\ACon{Traversal}} structure:
  % \LibCode\FTraversalIdProofRest
  \AppCode\FTraversal

@@

  This prevented us to define a useful lemma, which also needs to
  be instantiated separately for variables and terms: the counterpiece
  to \ACode{\AField{wk-id/`}}, which we call \ACode{\AField{wk-`/id}}.

@@

  Using the \ACode{\ACon{WkKit}}, we then define
  the \ACode{\ACon{ComposeKit}} and \ACode{\ACon{ComposeTraversal}}
  structures, which serve analogous purposes to \ACode{\ACon{Kit}}
  and \ACode{\ACon{Traversal}} from before:
  the library provides a \ACode{\ACon{ComposeKit}} for each of the
  four compositions, whereas the \ACode{\ACon{ComposeTraversal}}
  structure needs to be instantiated by the application with
  lemmas parameterized over arbitrary \ACode{\ACon{ComposeKit}}s.
  This gives the user a composition operation
  \ACode{\AField{\_·ₖ\_}} between arbitrary maps and corresponding
  lemmas, instead of multiple definitions and lemmas about concrete
  maps, enabling us to reason on a more abstract level and to avoid
  further boilerplate.

@@

  Similar as for the \ACode{\ACon{Traversal}}, we derive further
  definitions from the the axiom of a \ACode{\ACon{ComposeTraversal}}:
  \begin{itemize}
  \item 
    A map \ACode{ϕ} followed by a weakening is equivalent to a
    weakening followed by \ACode{ϕ} that has been lifted over the weakened
    variable:
    \LibCode\KCommLiftWeaken
    % \LibCode\KCommLiftWeakenProof
    By applying \ACode{\AField{⋯-fusion}} to this lemma, we get the same
    statement for multiple map applications:
    \LibCode\KCommLiftWeakenTraverse
    % \LibCode\KCommLiftWeakenTraverseProof
  \item 
    A weakening followed by a singleton substitution cancel each other out:
    \LibCode\KWeakenCancelsSingle
    % \LibCode\KWeakenCancelsSingleProof
    By applying \ACode{\AField{⋯-fusion}} and \ACode{\AField{⋯-id}} to
    this lemma, we get the same statement for multiple map applications:
    \LibCode\KWeakenCancelsSingleTraverse
    % \LibCode\KWeakenCancelsSingleTraverseProof
  \item 
    A singleton substitution followed by some map \ACode{ϕ} is the same
    as \ACode{ϕ} lifted over the variable of the singleton substitution
    followed by the singleton substitution with \ACode{ϕ} applied to
    its content:
    \LibCode\KDistLiftSingle
    % \LibCode\KDistLiftSingleProof
    By applying \ACode{\AField{⋯-fusion}} to this lemma, we get the
    same statement for multiple map applications:
    \LibCode\KDistLiftSingleTraverse
    % \LibCode\KDistLiftSingleTraverseProof
  \item 
    Given the \ACode{\ADef{⋯-↑-wk}} lemma, we can define the four
    \ACode{\ACon{ComposeKit}} instances:
    \LibCode\KComposeKitInstances
    \LibCode\KComposeKitInstancesConcrete
  \end{itemize}

  \subsection{Instantiation for System F}
  In this subsection, we show how to instantiate the
  \ACode{\ACon{ComposeTraversal}} abstraction for System F.
  In practice, this is done by our reflection algorithm automatically,
  but it can be instructive to see what happens under the hood.

  \AppCode\FFusion
  \AppCode\FFusionProofInteresting
  The interesting cases are:
  \begin{itemize}
  \item 
  \end{itemize}

  % \AppCode\FFusionProofRest
  \AppCode\FComposeTraversal

@@

  In the context of multi-sorted syntax, the notation of a type can be
  described as a mapping between sorts.
  For System F, the expression sort \ACode{\ACon{𝕖}} maps to the type sort
  \ACode{\ACon{𝕥}} and the sort \ACode{\ACon{𝕥}} maps to the kind sort
  \ACode{\ACon{𝕜}}.
  To teach our framework about types, we have to instantiate the following structure:
  \LibCode\KTypeSorts
  For System F the instantiation is
  \AppCode\FTypes
  For a nicer interface, we define a type analogously to a term:
  \LibCode\KTypes

@@

  \section{Typing}
  \label{sec:typing}
  Equipped with a notion of types and type context we are ready to define
  the multi-sorted typing relation for System F, incorporating both
  typing and kinding:
  \AppCode\FTyping
  The interesting cases are:
  \begin{itemize}
  \item
    the variable typing constructor covers both expression- and
    type-variables, analogously to the variable term constructor;
  \item
    the typing of the lambda abstraction \ACode{\ACon{⊢λ}} weakens the
    type of the codomain \ACode{t₂}.
    This is necessary, because with multi-sorted syntax, types could
    also depend on expressions, so the sub typing derivation accounts for
    another variable which is not used by the type.
  \item
    the kinding constructor \ACode{\ACon{⊢τ}} states that all types
    have kind \ACode{\ACon{★}}. This is sufficient for System F as types
    are automatically well-scoped due to intrinsic scoping.
  \end{itemize}
  To teach the framework about typing, we create a structure analogously to
  \ACode{\ACon{Syntax}}:
  \LibCode\KTyping
  and instantiate it for System F:
  \AppCode\FTypingInst

@@

  Given a \ACode{\ACon{TypingTraversal}}, we can easily instantiate the
  \ACode{\ACon{TypingKits}} for variables and terms:
  \LibCode\KTypingInstances
  % \LibCode\KTypingTraversalNotation

  \subsection{Instantiation for System F}
  In this subsection, we show how to instantiate the
  \ACode{\ACon{TypingTraversal}} abstraction for System F.
  This is the only structure that is \emph{not} instantiated
  automatically via reflection, as typing relations can be arbitrary
  complex in general.

  \AppCode\FPreserve
  The interesting parts of the proof are:
  \begin{itemize}
  \item 
  \end{itemize}

  \AppCode\FTypingTraversal

  % As the \ACode{\ACon{Traversal}} and  \ACode{\ACon{ComposeTraversal}} instances
  % can be generally derived via reflection,
  % the \ACode{\AField{\_⊢⋯\_}} lemma is in fact the only lemma that
  % has to been proved manually for subject reduction.
  % From this single lemma follows type preservation for all operations
  % of both renamings and substitutions for all sorts of variables
  % simultaneously.

@@

  \section{Reflection \& Generics}
  \label{sec:reflection}
  We used Agda's reflection mechanism\cite{TODO} to derive
  instantiations related to all structures for untyped substitition,
  i.e. \ACode{\ACon{Traversal}} and \ACode{\ACon{ComposeTraversal}}.

  To gain insight into the class of object languages supported by our
  reflection algorithm, we modeled a class of object languages using a
  generic programming technique and instantiated the structures
  generically. The class contains all object languages which:
  \begin{enumerate}
  \item Have a variable constructor of type TODO;
  \item Uses subterms only directly (e.g. not in lists);
  \item Subterms only extend the scope-context.
  \end{enumerate}

@@

  \section{Case Studies}
  \label{sec:casestudies}
  \begin{itemize}
  \item
    For lambda calculus with dependent function types, the framework
    just works out of the box for both deriving and instantiating the
    \ACode{\ACon{TypingTraversal}}. Additionally, having the
    \ACode{\ACon{ComposeKit}} abstraction turned out to be useful
    to prove confluence, allowing to unify certain lemmas, which
    otherwise need to be proved for both renamings and substitutions,
    e.g. that for two terms \ACode{t} and \ACode{t'}
    with \ACode{t \ACon{↪} t'} it follows that \ACode{(t \AField{⋯} ϕ)
    \ACon{↪} (t' \AField{⋯} ϕ)} for any map \ACode{ϕ}.
  \item
    For System F with subtyping the main challenges are how to represent
    subtyping constraints and how to deal with the fact that
    substitution preserves typing is not generally true, as type variables
    have subtyping bounds that need to be respected.
    Instead of defining typing contexts and type preservation lemmas independent
    of our framework, we found a way that still allows us to benefit
    from the typing machinery.
    When binding a type variable with a subtyping constraint \ACode{α <: t},
    we first bind the type variable as \ACode{α : ★}, and then bind the constraint
    with a dummy name \ACode{\_ ∶ (α <: t)}. This description is
    similar to first-class constraints, but restricted enough to be
    isomorphic to the original formalization.
    With that encoding substitution preserves typing is generally true again, enabling
    the use of the framework's typing contexts and \ACode{\ACon{TypingKit}} machinery.
    Replacing a type variable \ACode{α <: t} with a type \ACode{t'}
    which is not a subtype of \ACode{t'}, the resulting term is still
    well-typed, but in a context with an insatisfiable constraint.
  \item
    Object language with pattern matching pose an interesting problem.
    A clause \ACode{p ⇒ e} in a match-expression consists of a pattern
    \ACode{p} and a continuation expression \ACode{e}.
    The pattern \ACode{p} can bind arbitrary many variables in subpatterns,
    which are then available as free variables in \ACode{e}.
    % The challenge is to track the sorts of variables bound by a pattern,
    % to make them available to \ACode{e} by extending \ACode{e}'s sort
    % context correspondingly.
    To track the sorts of variables bound by a pattern, one can define
    the sort of patterns to contain the sorts of the variables that it binds:
    % 𝕡 : List (Sort Var) → Sort NoVar
    The syntax for variable patterns and clauses can then be specified as:
    % varpat : S ⊢ 𝕡 [ 𝕖 ]  -- pattern binding an expression variable
    % pairpat : S ⊢ 𝕡 S₁ → S ⊢ 𝕡 S₂ → S ⊢ 𝕡 (S₁ ++ S₂)
    % clause : S ⊢ 𝕡 S' → (S' ++ S) ⊢ 𝕖 → S ⊢ 𝕔
  \end{itemize}

@@

  Allais et al.\cite{DBLP:conf/cpp/Allais0MM17} propose a powerful
  abstraction for denotational semantics and semantic fusion lemmas.
  In later work\cite{DBLP:journals/pacmpl/AllaisA0MM18}, they use
  generic programming to instantiate this abstraction for a class of
  object languages comparable to ours.
  They demonstrate how both renamings and substitutions can be
  described as semantics and how the four composition lemmas follow from
  their generic fusion lemma.
  They do cover an abstraction to unify renamings and substitutions.
  They do not show how to work with multiple variable sorts, how to
  abstract over the compositions, and how to abstract over typing in
  an extrinsically typed scenario.
  However, we believe that their framework is general enough such that
  our concepts can also be transferred to their framework.
